---
title: "HW02 - Basic data types and vectors"
subtitle: "Stat 133, Fall 2016, Prof. Sanchez"
author: "Your Name"
output: html_document
---

The purpose of this assignment is to work with vectors of different
data types: numeric (integer and real), character, logical, as well as factors.

Download this file or open an `Rmd` file in RStudio and copy the
contents of this file.


## Data

The data for this assignment is in the `cereals.RData` file, 
located in the github repository (this file is in binary format). 

Start a new session in RStudio, and make sure there's no saved objects:
```{r eval = FALSE}
rm(list = ls())
```

Use the following code to read the data from the github repository:
```r
# download RData file into your working directory
rdata <- "https://github.com/ucb-stat133/stat133-fall-2016/raw/master/data/cereals.RData"
download.file(url = rdata, destfile = 'cereals.RData')

# load data in your R session
load('cereals.RData')
```

Once you've read the data, list the objetcs with `ls()`.
You should be able to see five objects:

- `cereals`
- `brands`
- `calories`
- `sugars`
- `shelfs`

Check the class of the objects and make sure they are vectors. Then, verify 
they all have the same length.
```{r}
# your code here

```

Convert `brands` and `shelfs` into factors.
```{r}
# your code here

```


### Exploring Quantitative Variables

For vectors `calories` and `sugars` plot:

- histograms
- boxplots (horizontal orientation)
- get a numeric summary with `summary()`

```{r}
# your code here

```


### Exploring Qualitative Variables

For factors `brands` and `shelfs`:

- Find how many unique brands
- Find how many different types of shelfs
- Get frequencies with `table()`
- plot a barchart
- plot a piechart
- plot a dotchart

```{r}
# how many unique brands


# frequencies of brands


# barchart (freqs of brands)


# piechart (freqs of brands)


# dotchart (freqs of brands)


#
```


```{r}
# how many different types of shelfs


# frequencies of shelfs


# barchart (freqs of shelfs)


# piechart (freqs of shelfs)


# dotchart (freqs of shelfs)


#
```


### Two quantitative variables

- Calculate the correlation between `calories` and `sugars`
- Plot a scatter diagram of `sugars` (x-axis) and `calories` (y-axis)

```{r}
# your code here

```


### Two variables: Sugars and Brands

The following command creates a vector with the amount of sugars for those
cereals of Kellog's brand:
```{r eval = FALSE}
# sugars for Kellogs
sugars_kellogs <- sugars[brands == 'Kellogs']

```

- Write similar commands to get vectors of sugars for each brand:
```{r}
# vectors of sugars for other brands

```

- Now write R code to get the average amount of sugars for each brand:
```{r}
# avg sugars for other brands

```

- Create a vector called `sugars_brands` that contains the amount of sugars 
for each brand. This vector should have as many elements as number of unique 
brands.
```{r}
# your 'sugars_brands' vector

```


- Find how to use `boxplot()` to get a conditional plot:
`sugars` conditioned to `brands`. This graphic displays, in a single plot frame,
a boxplot (of sugars) for each type of brand.
```{r}
# your conditional boxplot

```

-----
