---
title: "HW03 - Stat 133, Fall 2016"
author:
- Prof. Sanchez
- LAB 201
output: html_document
urlcolor: blue
---

```{r echo = FALSE, message= FALSE}
# necessary packages
library(xtable)
```

The purpose of this assignment is to write simple functions, as well as 
practicing using conditionals and loops in R. Please turn in 
a physical copy of the knitted html file during lab, and submit both your 
`.Rmd` and `.html` files to bCourses. Due Friday Oct-07.

-----

## Area of a circle

For a given circle of radius $r$, the area $A$ is:
$$
A = \pi r^2
$$

Write a function `circle_area()` that calculates the area of a circle. This 
function must take one argument `radius`. Give `radius` a default value of 1.
The function should `stop()` if `radius` is negative. In the code chunk options,
please include `error = TRUE` so that you don't get an error when knitting 
the file.
```{r, error = TRUE}
# your circle_area() function

```

Test your function:
```{r error = TRUE}
# circle_area()

# circle_area(radius = 3)

# circle_area(radius = -2)
```


## Area of a cylinder

For a given cylinder of radius $r$ and height $h$ the area $A$ is:
$$
A = 2 \pi r h + 2 \pi r^2
$$

Notice that the formula of the area of a cylinder includes the area of a 
circle: $\pi r^2$. Write a function `cyl_area()`, that calls `circle_area()`, 
to compute the area of a cylinder. 

This function must take two arguments: `radius` and `height`. Give both 
arguments a default value of 1. In addition, the function should stop if any
of `radius` or `height` are negative.
```{r, error = TRUE}
# your cyl_area() function

```

Test your function
```{r error = TRUE}
# cyl_area()

# cyl_area(radius = 2, height = 3)

# cyl_area(radius = -2, height = 1)

# cyl_area(radius = 2, height = -1)

# cyl_area(radius = -2, height = -1)

```


## Volume of a cylinder

For a given cylinder of radius $r$ and height $h$ the volume $V$ is:
$$
V = \pi r^2 h
$$

Write a function `cyl_volume()`, that calls `circle_area()`, to compute the 
volume of a cylinder. This function must take two arguments: 
`radius` and `height`. Give both arguments a default value of 1.

```{r, error=TRUE}
# your cyl_volume() function

```

Test your function:
```{r error = TRUE} 
# cyl_volume()

# cyl_volume(radius = 3, height = 10)

# cyl_volume(height = 10, radius = 3)
```

-----

\newpage

## Currency Converter

Consider the exchange rates of one US dollar for the following currencies
(source: XE currency table US dollar, 09-20-2016):

| Currency         | name     | rate    |
|------------------|----------|--------:|
| US dollar        | `dollar` | 1.00    |
| Euro             | `euro`   | 0.89    |
| British pound    | `pound`  | 0.77    |
| Japanese yen     | `yen`    | 101.69  |
| Chinese yuan     | `yuan`   | 6.67    |
| South Korean Won | `won`    | 1118.21 |
| Indian rupee     | `rupee`  | 66.98   |
| Mexican peso     | `peso`   | 19.82   |
| Brazilian real   | `real`   | 3.25    |

Write a function `exchange()` that converts from one currency to another. 
This function takes three arguments:

- `amount` is a positive numeric input
- `from` is a character string indicating the name of a currency
- `to` is a character string indicating the name of a currency

Give these arguments default values of `amount = 1`, `from = "dollar"`, and 
`to = "euro"`. 
Inside `exchange()` you must declare a named vector with the given 
exchange rates:
```{r}
exchange <- function(amount = 1, from = "dollar", to = "euro") {
  # vector of dollar exchange rates
  x <- c(
    dollar = 1,
    euro = 0.89,
    pound = 0.77,
    yen = 101.69,
    yuan = 6.67,
    won = 1118.21,
    rupee = 66.98,
    peso = 19.82,
    real = 3.25)
  
  # write the rest of the code of your function
  # ...
}
```

- You can use any control flow structure: `if-else`, `switch()`, `for` loops, 
`while`, `repeat`.
- You can use any data structures inside `exchange()`: vectors, matrices, 
data frames, lists, etc.
- The output must be a numeric vector

Test your `exchange()` function with:
```{r error = TRUE}
# exchange()

# exchange(amount = 10, from = 'euro', to = 'peso')

# exchange(amount = 20, from = 'yuan', to = 'pound')

# exchange(amount = 30, from = 'rupee', to = 'won')
```

-----


## Two Given Points

Let $p_1$ and $p_2$ be two points with two coordinates: $p_1 = (x_1, y_1)$ and 
$p_2 = (x_2, y_2)$.

The distance $d$ between two points can be calculated with the formula:
$$
d = \sqrt{(x_2 - x_2)^2 + (y_2 - y_1)^2}
$$

The midpoint of the line segment between $p_1$ and $p_2$ can be found as:
$$
p = \left( \frac{x_1 + x_2}{2}, \frac{y_1 + y_2}{2} \right)
$$

The intercept $a$ and the slope $b$ of the line $y = a + bx$ connecting two 
points $p_1$ and $p_2$ can be found as:
$$
b = \frac{y_2 - y_1}{x_2 - x_1}, \quad a = y_1 - b x_1
$$


#### Distance

Write a function `find_distance()` that returns the distance between two given
points. You should be able to call the function like this:
```r
# coordinates for point-1 and point-2
p1 <- c(0, 0)
p2 <- c(1, 1)

find_distance(p1, p2)
```

Your function:
```{r, error = TRUE}
# your find_distance() function

```

Test your `find_distance()` function with:
```{r error = TRUE}
# NOTE: Uncomment the following lines and then knit this file to check your function is working as required

# p1 <- c(1, 10)
# p2 <- c(10, 20)
# find_distance(p1, p2)
```

#### Midpoint

Write a function `find_midpoint()` that returns the midpoint between two given 
points. You should be able to call the function like this:
```r
p1 <- c(0, 0)
p2 <- c(1, 1)

find_midpoint(p1, p2)
```

Your function:
```{r, error = TRUE}
# your find_midpoint() function

```

Test your `find_midpoint()` function with:
```{r error = TRUE}
# NOTE: Uncomment the following lines and then knit this file to check your function is working as required

# p1 <- c(1, 10)
# p2 <- c(10, 20)
# find_midpoint(p1, p2)
```


#### Slope

Write a function `find_slope()` that returns the slope of the line connecting 
two given points. You should be able to call the function like this:
```r
p1 <- c(0, 0)
p2 <- c(1, 1)

find_slope(p1, p2)
```

Your function:
```{r, error = TRUE}
# your find_slope() function

```

Test your `find_slope()` function with:
```{r error = TRUE}
# NOTE: Uncomment the following lines and then knit this file to check your function is working as required

# p1 <- c(1, 10)
# p2 <- c(10, 20)
# find_slope(p1, p2)
```


#### Intercept

Write a function `find_intercept()` that returns the intercept of the line 
connecting two given points. This function must internally use `find_slope()`
```r
p1 <- c(0, 0)
p2 <- c(1, 1)

find_intercept(p1, p2)
```

Your function:
```{r, error = TRUE}
# your find_intercept() function

```

Test your `find_intercept()` function with:
```{r error = TRUE}
# NOTE: Uncomment the following lines and then knit this file to check your function is working as required

# p1 <- c(1, 10)
# p2 <- c(10, 20)
# find_intercept(p1, p2)
```

#### Line

Write a function `find_line()`. This function must use `find_slope()` and
`find_intercept()`. The output should be a list with two named elements:
`"intercept"` and `"slope"`, Here is how you should be able to use 
`find_line()`:
```r
p1 <- c(0, 0)
p2 <- c(1, 1)

eq <- find_line(p1, p2)
eq$intercept
eq$slope
```

Your function:
```{r, error = TRUE}
# your find_line() function

```

Test your `find_line()` function with:
```{r error = TRUE}
# NOTE: Uncomment the following lines and then knit this file to check your function is working as required

# p1 <- c(1, 10)
# p2 <- c(10, 20)
# eq <- find_line(p1, p2)
# eq$intercept
# eq$slope
```

#### Information about two given points

Once you have the functions `find_distance()`, `find_midpoint()`, and 
`find_line()`, write an overall function called `info_points()` that returns a 
list with the distance, the midpoint, and the line's slope and intercept terms. 
Here is how you should be able to use `info_points()`:

```r
p1 <- c(-2, 4)
p2 <- c(1, 2)

results <- info_points(p1, p2)
results$distance
results$midpoint
results$intercept
results$slope
```

Your function:
```{r, error = TRUE}
# your info_points() function

```

Test your `find_line()` function with:
```{r error = TRUE}
# NOTE: Uncomment the following lines and then knit this file to check your function is working as required

# p1 <- c(1, 10)
# p2 <- c(10, 20)

# results <- info_points(p1, p2)
# results$distance
# results$midpoint
# results$intercept
# results$slope
```

Use the following code to create a plot that displays the given points, 
the line, and the midpoint. Note that the title of the plot shows the line 
equation.
For instance, if the points are $p_1 = (-2, 4)$ and $p_2 = (1, 2)$, the plot 
may look like this (you should choose different points!):

```{r}
# change these points and pass them to info_points()
p1 <- c(-2, 4)
p2 <- c(1, 2)

plot.new()
# depending on your chosen points you may have to set different limits
plot.window(xlim = c(-3, 3), ylim = c(0, 5))
axis(side = 1)
axis(side = 2, las = 1)
points(p1[1], p1[2], cex = 1.5, col = "#FF8834", pch = 19)
points(p2[1], p2[2], cex = 1.5, col = "#FF8834", pch = 19)
# midpoint (here you should use the midpoint outputs of your function)
points(-1/2, 3, cex = 1.5, pch = "x", col = "#E16868")  
# slope and intercept (here you should use the outputs of your function)
abline(a = 8/3, b = -2/3, col = "#FF883477", lwd = 3) 
title(main = expression(paste(y, ' = ', (-2/3) * x, ' + ', (8/3))))
```