---
title: "HW03 - Data tables and graphics"
subtitle: "Stat 133, Fall 2016, Prof. Sanchez"
author: "Your Name"
output: html_document
---

This assignment has 4 main purposes:

- practicing reading data tables
- working with data frames, vectors and factors
- practice basic manipulation of data structures
- creation of simple graphics with `"ggplot2"`

Download this file or open an `Rmd` file in RStudio and copy the
contents of this file.


## Data set Camping Tents

Start by practicing reading tables in R. The data for this assignment is in the 
file `camping-tents.csv` (available in the folder `data/` in the 
github repository of the course)

Use `download.file()` to download the file `camping-tents.csv` in your computer.
You'll have to specify your own destination (`destfile`)
e.g. in my case I'm specifying: `"~/Downloads/camping-tents.csv"`
(your `destfile` will probably be different!)
```{r eval = FALSE}
download.file(
  url = 'https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2016/master/data/camping-tents.csv', 
  destfile = '~/Downloads/camping-tents.csv')
```

- __Importing A:__ Use `read.csv()` to import `camping-tents.csv` and assign 
it to an object called `tdf1`
```{r}
# import using 'read.csv()'

```


- __Importing B:__ Use `read.csv()` to import `camping-tents.csv` and assign 
it to an object called `tdf2`. In this case, strings must NOT be converted into 
factors!
```{r}
# import using 'read.csv()' without converting strings into factors

```


- __Importing Option 3:__ Create a character vector `col_class` with the 
following classes for the columns (to be used with argument `colClasses`)
  - name:     `character`
  - brand:    `character`
  - price:    `numeric`
  - weight:   `numeric`
  - height:   `integer`
  - bestuse:  `factor`
  - seasons:  `factor`
  - capacity: `factor`

Use `read.table()` to import `camping-tents.csv` and assign it to 
an object called `tdf3`. Pass the vector `col_class` to the argument 
`colClasses`, and specify the right field separator, and header.

```{r}
# import using 'read.csv()' specifying `colClasses

```


- __Importing Option 4:__ Use `read_csv()` from package `"readr"` to import 
`camping-tents.csv` and assign it to an object called `tents`.
Use the argument `col_types` to specify column types:
  - name:     `character`
  - brand:    `character`
  - price:    `numeric`
  - weight:   `numeric`
  - height:   `integer`
  - bestuse:  `factor`
  - seasons:  `factor`
  - capacity: `factor`


```{r}
# import using 'read_csv()'

```


-----

## Inspecting data frames

Once you have your data frame `tents`, the next step 
involves performing a first exploration.

```{r}
# inspect the data structure of 'tents'


# check the class of the object 'tents'


# how many rows in the dataset?


# how many columns in the dataset?


# names of columns


# take a look at the first 5 rows


# take a look at the last 3 rows

```


### Quantitative Variables:

Exploring the quantitative variables

1. `price`
2. `weight`
3. `height`

Instead of using the `summary()` function, find functions
that allow you to get the following summaries for `price`

- minimum value
- first quartile (i.e. 25th percentile)
- third quartile (i.e. 75th percentile)
- 90th percentile
- median value (i.e. 50th percentile)
- mean value
- maximum value
- inter-quartile range (IQR)
- standard deviation

```{r}
# your code

```


Add a new variable `weight_lbs` to the data frame `tents` for weight in pounds.
Keep in mind that `weight` is given in grams.
```{r}
# add weight_lbs

```

Add a new variable `height_in` to the data frame `tents` for height in inches.
Keep in mind that `height` is given in centimeters.
```{r}
# add height_in

```

Write R code (using bracket notation) to answer the following questions:
```{r}
# how many tents have a price greater than $400


# how many tents have a price less than or equal to $300


# how many tents have a price between $300 and $400
# (including both $300 and $400 prices)



# what's the name of the tent with maximum price


# what's the name of the tent with minimum price


# what's the name of the tent with maximum weight


# what's the name of the tent with minimum weight


# select the data of tents with 
# price > $400 AND weight < 1500 grams




# calculate the 90th percentile for height and 
# assign it to the object 'height_p90'


# calculate the 90th percentile for weight and 
# assign it to the object 'weight_p90'


# select the data of tents with 
# height > height_p90 AND weight > weight_p90

```


-----

## More Manipulations

Use `subset()` to obtain a new data frame called `rei` containing the data of 
those tents with brand `'rei'`; and inspect its dimension:
```{r}
# rei data frame

```


Use the `cut()` function to create a factor `price_cut` from all the tent 
prices; for the argument `breaks` give a vector of cutting points
such that the obtained levels are as follows:
```
  (0, 100]
(100, 200]
(200, 300]
(300, 400]
(400, 500]
(500, 600]
(600, 700]
```

```{r}
# your price_cut factor

```

Use `table()` to check the obtained frequencies of `price_cut`
```{r}
# confirm your frequencies with `table()`

```


Use `cut()` to create a factor `weight_cut` by using the
breaking points and labels according to the following table:
```
-----------------------
     intervals | labels
---------------|-------
     (0, 1000] |  1kg
  (1000, 2000] |  2kg
  (2000, 3000] |  3kg
  (3000, 4000] |  4kg
  (4000, 5000] |  5kg
  (5000, 6000] |  6kg
  (6000, 7000] |  7kg
  (7000, 8000] |  8kg
  (8000, 9000] |  9kg
----------------------
```

```{r}
# your weight_cut factor

```

Verify that the frequencies given by `table(weight_cut)` are:
```
 1kg 2kg 3kg 4kg 5kg 6kg 7kg 8kg 9kg 
   3  24  30   7   7   2   9   3   3 
```

```{r}
# your verification

```


-----

## Statistical graphics of Quantitative Variables

Use `"ggplot2"` to obtain the following plots for each quantitative variable: 
(look at each graphic carefully and see what types 
of distribution patterns show each variable)

- histogram
- boxplot
- density curve

```{r}
# your plots

```


### Scatter diagrams

Obtain scatter plots of:

- price, height
- price, weight
- height, weight

```{r}
# your scatter diagrams

```



### Plot challenge

Here's a small challenge: Obtain a scatter plot of `weight` and `height` 
such that:

- label of x-axis is: `Weight (in grams)`
- label of y-axis is: `Height (in csm)`
- Points are colored with alpha transparency `alpha = 0.7`
- The symbol of points are filled triangles
- x-axis ranges from 0 to 10000
- y-axis ranges from 80 to 220
- Include a title

```{r}
# your scatter plot

```


-----

## Statistical graphics of Qualitative Variables

Now let's focus on the qualitative variables:

1. `brand`
2. `bestuse`
3. `seasons`
4. `capacity`

```{r}
# get frequency tables of each qualitative variable


# what is the brand with less number of tents


# are there any tents of brand 'rei'?
# and if so, how many?


# are there any tents of brand 'mountain-hardwear'
# and if so, how many?


# how many 'rei' tents are intended to be
# used ('bestuse') for Mountaineering?


# what unique brands have tents intended to be used for 'Mountaineering'


```


### Plot challenge: 

Make a barplot for 'brand' such that:

- bars are horizontally oriented
- bars are arranged in decreasing order
- brand labels are perpendicular to the y-axis
  (i.e. labels horizontally oriented)
- modify margins in order to have enough room for labels 
- x-axis ranges from 0 to 25
- include a title

```{r}
# your bar-chart

```


### Bivariate Plots

Use `geom_boxplot()` to plot the distribution of `weight` conditional to:
(get a boxplot for each categorical variable)

- brand
- bestuse
- capacity
- seasons

```{r}
# your boxplots

```


Make a scatter-plot of `height` and `weight`, using 
`capacity` as a factor for the color argument
```{r}
# scatter-plot of `height` and `weight` (color `capacity`)

```

Get the same scatter plot as above but now pass `bestuse`
as a factor for the color argument
```{r}
# scatter-plot of `height` and `weight` (color `bestuse`)

```



### Plot challenge

Obtain a new data frame called `big_agnes` by subsetting tents of brand 
`'big-agnes'`. Then, use this data frame to make a scatter plot of `weight` 
and `height`:

- The background of the entire plot must be white
- Instead of points, show the name of each big-agnes tent
- Set the color attribute based on `bestuse`
- Include a title

```{r}
# scatter-plot using `big_agnes` data frame

```


