---
title: "Stat 133, Fall 2016, More XML Practice"
author: "Gaston Sanchez"
date: "November 18, 2016"
output: html_document
---


### Parsing XML content

Load the package "XML"

```{r}
library(XML)
```

Get a copy of the _Faculty_ page of the Department of Statistics
```{r, eval=FALSE}
# faculty stats department
url <- 'http://statistics.berkeley.edu/people/faculty'

download.file(url, 'faculty.html')
```

Inspect the HTML source in your browser

Let's parse the html document with `htmlParse()`:
```{r}
# parse html content
# (output as an internal C-structure)
doc <- htmlParse('~/Documents/faculty.html')

class(doc)
```

Identify the node for each professor. Get the following content:

- Name
- Job title
- Email
- Office
- website link

-----

Getting the Names
```{r}
name_nodes <- getNodeSet(doc, '//div[@class="views-field views-field-title"]')

sapply(name_nodes, xmlValue)                       
```

Alternatively
```{r}
xpathSApply(doc, '//div[@class="views-field views-field-title"]', xmlValue)
```

-----

Getting the Title
```{r}
title_nodes <- getNodeSet(doc, '//div[@class="views-field views-field-field-job-title"]')

sapply(title_nodes, xmlValue)                       
```

Alternatively
```{r}
xpathSApply(
  doc, 
  '//div[@class="views-field views-field-field-job-title"]',
  xmlValue)
```

-----

Getting the email
```{r}
email_nodes <- getNodeSet(doc, '//div[@class="views-field views-field-field-email"]')

sapply(email_nodes, xmlValue)                       
```

Alternatively
```{r}
xpathSApply(
  doc, 
  '//div[@class="views-field views-field-field-email"]',
  xmlValue)
```

-----

Getting the office
```{r}
office_nodes <- getNodeSet(
  doc, 
  '//div[@class="views-field views-field-field-location""]')

sapply(office_nodes, xmlValue)                       
```

Alternatively
```{r}
xpathSApply(
  doc, 
  '//div[@class="views-field views-field-field-location"]',
  xmlValue)
```

-----

Getting the links
```{r}
link_nodes <- getNodeSet(
  doc, 
  '//div[@class="views-field views-field-title"]/h2/a')

sapply(link_nodes, xmlAttrs)
```

Alternatively
```{r}
xpathSApply(
  doc, 
  '//div[@class="views-field views-field-title"]/h2/a',
  xmlAttrs)
```



-----

## Courses in the Department of Statistics

Get the courses for the fall 2015
[http://statistics.berkeley.edu/courses/fall-2015](http://statistics.berkeley.edu/courses/fall-2015)

- Name
- Section
- Days and Time
- Room
- Instructor
