---
title: "Stat 133, Fall 2016, Manipulating Strings"
author: "Gaston Sanchez"
date: "October 3, 2016"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes: \usepackage{float}
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## Strings Manipulation

The purpose of this lab is to work with regular expressions and string 
manipulations in R. You will need to work with the R packages `"XML"` and
`"stringr"`. 
```{r, message=FALSE, warning=FALSE}
library(XML)
library(stringr)
```


## Data 

You will be working with the data set for the _Men's discuss throw world record
progression_, available in wikipedia:

[https://en.wikipedia.org/wiki/Men%27s_discus_throw_world_record_progression](https://en.wikipedia.org/wiki/Men%27s_discus_throw_world_record_progression)


Let's start by downloading the html file of the wikipedia page. To do this, 
we will use the function `paste0()` to assemble the address of the webpage.
We can create a base url `wiki`, and then a separate string with the 
part that has to do with the Men's discus throw world records: 
```{r eval=FALSE}
# Download a copy of the HTML file, 
wiki = "https://en.wikipedia.org/wiki/"
men_discus = "Men%27s_discus_throw_world_record_progression"

wiki_men_discus = paste0(wiki, men_discus)

download.file(wiki_men_discus, "men-discus-throw.html")
```

Once you've downloaded the html content in the file `men-discus-throw.html`, 
you can use the function `readHTMLTable()` (from `"XML"`) to read the tables in 
the html document. The function `readHTMLTable()` parses an html document and 
extracts all html tables. The output is an R list with
as many data frames as html tables present in the source document:
```{r}
# read in tables with readHTMLTable()
tbls <- readHTMLTable("men-discus-throw.html")

# how many html tables?
length(tbls)
# dimension of tables?
lapply(tbls, dim)
```

As ou can tell, `tbls` contains `r length(tbls)` HTML tables. Inspecting their 
dimensions, you should see that it is the first table the one that we are 
interested in. So let's re-read just the first table, and tell R to not convert
strings as factors:
```{r}
# the first table is the good one
dat <- readHTMLTable("men-discus-throw.html", which = 1,
                     stringsAsFactors = FALSE)
```

Use `head()` and/or `str()` to look at the content and inspect the structure
of the data frame `dat`.


## Data Cleaning

The main goal is to clean the downloaded data frame in order to produce another 
data frame with columns:

- `mark` (in meters)
- `first_name` (first name of athlete)
- `last_name` (last name of athlete)
- `initials` (name initials)
- `day` (number of day)
- `month` (name of month)
- `year` (number of year)


## Regex 

Here's a table with some of the common regex patterns:

| Pattern | Description   |
|---------|---------------|
| `abc`   | letters       |
| `123`   | digits        |
| `.`     | any chracter  |
| `\\.`   | period        |
| `[abd]` | only a, b, or c |
| `[^abc]` | not a, b, nor c |
| `[a-z]` | characters a to z |
| `[0-9]` | numbers 0 to 9 |
| `{m}`   | _m_ repetitions |
| `{m,n}` | _m_ to _n_ repetitions |
| `*`     | zero or more repetitions |
| `+`     | one or more repetitions |
| `?`     | optional character |
| `\\d`   | any digit     |
| `\\D`   | any Non-digit |
| `\\w`   | any alphanumeric character |
| `\\W`   | any non-alphanumeric character |
| `\\s`   | any whitespace |
| `\\S`   | any Non-whitespace |
| `^a`    | starts with _a_ |
| `b$`    | ends with _b_ |




## Extracting Mark (in meters)

The column `Mark` contains a character string with the record expressed 
both in meters and feet-inches. We want to extract only the value associated
to meters.

My suggestion is to always start small. In this case, we can get a subset of
values on which we can play with:
```{r}
tmp <- head(dat$Mark)
tmp
```

With the values in `tmp`, let's try to match the numeric values of meters.

__Meters: Option 1.__
One possibility is to use `str_split()` to __split__ the vector using `"m"` as 
the pattern to separate the values.
```{r}
str_split(tmp, "m")
```
The output is a list in which each element has two values: the character 
numbers of the meters, and the characters inside parenthesis of feet-inches.

Then we can use `sapply()` to loop over the list, and retrieve the first 
element:
```{r}
tmp_split <- str_split(tmp, "m")
sapply(tmp_split, function(x) x[1])
```

Finally, we need to remove the extra spaces at the end of each string. One
option to do this is with `str_trim()`. The last step consists in converting
the characters into a numeric vector:
```{r}
# removing extra white spaces
tmp_meters <- sapply(tmp_split, function(x) x[1])
str_trim(tmp_meters)

# converting to numeric
as.numeric(str_trim(tmp_meters))
```


__Meters: Option 2.__ Another alternative to extract the height value of meters,
is to use a regular expression that matches those values. 
We need a pattern that matches two digits, followed by a dot `"."`, followed
by two more digits.

Let's start by matching those marks that start with 4 (e.g. 47.58 47.61 47.89)
```{r}
# matching the first 4
str_extract(tmp, "^4")
```
Note the use of the caret symbol `"^"` to indicate that we are matching the
character `"4"` at the beginning of the string.

Likewise, we can specify a character `"5"` to match marks starting with 5""
```{r}
# matching the first 5
str_extract(tmp, "^5")
```

You can use a character class `[0-9]` which represents any digit:
```{r}
# matching the first digit
str_extract(tmp, "^[0-9]")
```

Or you can be more precise about a certain range of digits, for example, from
4 to 7:
```{r}
# matching the first digit
str_extract(tmp, "^[4-7]")
```

If you want to match any two digits (one next to the other), then combine two 
digit classes:
```{r}
# matching the first two digits
str_extract(tmp, "^[0-9][0-9]")
```

To match the literal dot `"."`, you need to escape it in R using double
backslashes:
```{r}
# height in meters
str_extract(tmp, "^[0-9][0-9]\\.[0-9][0-9]")
```

The pattern `"^[0-9][0-9]\\.[0-9][0-9]"` indicates: match a digit, followed by 
another digit, followed by a dot `\\.`, followed by another digit, followed by 
another digit.

There is some repetition in the pattern `"^[0-9][0-9]\\.[0-9][0-9]"`.
Try to simplify it using the quantifiers characters:

| Pattern | Description   |
|---------|---------------|
| `a{m}`   | _m_ repetitions of `"a"` |
| `a{m,n}` | _m_ to _n_ repetitions of `"a"` |
| `a*`     | zero or more repetitions of `"a"` |
| `a+`     | one or more repetitions of `"a"` |
| `a?`     | optional character `"a"` |

Try `str_extract()` with the following patterns:

- `"^[0-9]{1}"`
- `"^[0-9]{1,2}"`
- `"^[0-9]*"`
- `"^[0-9]+"`
- `"^[0-9]?"`

```{r}
# "^[0-9]{1}"


# "^[0-9]{1,2}"


# "^[0-9]*"


# "^[0-9]+"


# "^[0-9]?"

```

How would you simplify the pattern `"^[0-9][0-9]\\.[0-9][0-9]"`?
```{r}
# use quantifiers to simplify "^[0-9][0-9]\\.[0-9][0-9]"

```

Now try matching using the digit character `"\\d+"` (i.e. one or more 
numeric characters)::
```{r}
# use quantifiers and "\\d+" to simplify "^[0-9][0-9]\\.[0-9][0-9]"

```

Once you have a simple pattern, use it on the entire column `Mark` and get a
numeric vector:
```{r}
# numeric marks (in meters)
```


## Extracting Athlete Name

The second task involves extracting the first and last names of the athletes. 
If you inspect the column `Athlete`, you will see that all its values are 
formed with the first name, the last name, and the country inside parenthesis:
```{r}
ath <- head(dat$Athlete)
ath
```

Work with the sample vector `ath` and try to `str_extract()` the first name.
You can experiment with the _word_ pattern `"\\w+"` (i.e. one or more 
alphanumeric characters):
```{r}
# your code (for first name)

```

Now use the patterns _whitespace_ `"\\s"` and _word_ `"\\w+"` to attempt 
extracting the athlete's last name:
```{r}
# your code for the last name

```

Once you are done working with `ath`, use your code to extract the first and 
last names of all athletes; use vectors `first_name` and `last_name` for this
purpose:
```{r}
# first and alst name of all athletes
```


### Athlete's Initials

Use `first_name` and `last_name` to select the first letter in each vector
in order to form a new vector `initials` containing the initials of each 
athlete's name: e.g. `"J.T.", "T.L.". "G.H.", ...`
```{r}
# forming the initials vector
```


## Date

The date values are in the column `Date`:
```{r}
dts <- head(dat$Date)
dts
```


__Your turn:__  With the `dts` vector, extract in separate vectors the 
values of day, month name, and year: you can try using patterns such as 
`"[0-9]"`, `"\\d+"`, `"\\w+"`:
```{r}
# your code for days
```

```{r}
# your code for months
```

```{r}
# your code for years
```


## Clean Data Frame

Assuming that you have created vectors for all the cleaned components, you
should be able to create a data frame `discus`:

```{r echo = FALSE}
discus <- data.frame(
  # fill the data frame
)
```


## Exporting Clean Data

Use one of the writing table functions (e.g. `read.table()`) to export
the data frame `discus` to an external file (e.g. CSV file)

```{r}
# your code to save discus in a field-separated format file
```


## Extra challenge

Here's one challenge for you: write code (using a for-loop) to obtain a 
list `names_month` containing the unique athlete names in each month. 
Here's what `names_month` should look like for the first three elements:

```
$January
[1] "Edmund Piątkowski"

$March
[1] "John van"

$April
[1] "Bud Houser"       "Eric Krenz"       "Willy Schröder"   "Adolfo Consolini"
[5] "Al Oerter"        "Mac Wilkins"     
```


### Solutions at the end of this document



\newpage

# Solutions

```{r results='hide'}
# trying different quantifiers
str_extract(tmp, "^[0-9]{1}")
str_extract(tmp, "^[0-9]{1,2}")
str_extract(tmp, "^[0-9]*")
str_extract(tmp, "^[0-9]+")
str_extract(tmp, "^[0-9]?")
```

```{r}
# several ways to get numeric vector of mark (in meters)
mark <- as.numeric(str_extract(dat$Mark, "^[0-9][0-9]\\.[0-9][0-9]"))
mark <- as.numeric(str_extract(dat$Mark, "^[0-9]+\\.[0-9]+"))
mark <- as.numeric(str_extract(dat$Mark, "^\\d+\\.\\d+"))

# athlete first name
first_name = str_extract(dat$Athlete, "^\\w+")

# athlete last name
last_str <- str_extract(dat$Athlete, "\\s\\w+")
last_name <- str_trim(last_str)

# initials
first_initial <- str_extract(first_name, "\\w")
last_initial <- str_extract(last_name, "\\w")
initials <- paste0(first_initial, ".", last_initial, ".")

# extracting day
day = as.numeric(str_extract(dat$Date, "^[0-9]+"))

# extracting month
month = str_trim(str_extract(dat$Date, "\\D+"))

# extracting year
year = as.numeric(str_extract(dat$Date, "\\d+$"))
```


```{r}
# clean data frame
discus <- data.frame(
  mark = mark,
  first_name = first_name,
  last_name = last_name,
  initials = initials,
  day = day,
  month = month,
  year = year
)
```

```{r eval = FALSE}
# export data discus
write.csv(discus, file = "mean-discus-throw.csv", row.names = FALSE)
```

```{r}
# get unique month names
unique_months <- month.name[month.name %in% unique(month)]

# initialize names_month
names_month <- vector("list", length = length(unique_months))

# for loop to fill in names_month
for (i in seq_along(unique_months)) {
  first <- first_name[month == unique_months[i]]
  last <- last_name[month == unique_months[i]]
  names_month[[i]] <- unique(paste(first, last))
}
names(names_month) <- unique_months
```