---
title: "Lab 4: Intro to shiny documents"
subtitle: "Stat 133, Fall 2016"
author: "Gaston Sanchez"
output: html_document
---


> ### Learning Objectives
>
> - First contact with the R package "shiny"
> - What is shiny?
> - Understanding the concept of reactive elements
> - Creating Dynamic Plots


We'll work with the [babynames](https://github.com/hadley/babynames) data to 
visualize how your name (or other names) have changed over time in the USA.

Install the R package `babynames` (by Hadley Wickham) either from CRAN or 
from its github repo:
```r
# you may need to install "devtools""
install.packages("devtools")
library(devtools)

# install "babynames"
install_github("babynames", "hadley")
```

Now load these packages:
```{r, message=FALSE}
library(babynames)
library(ggplot2)
```

The data set `babynames` has `r nrow(babynames)` rows and `r ncol(babynames)`
columns. Column names are:

- `year`: calendar year (from 1880 to 2014)
- `sex`: sex of name
- `name`: name 
- `n`: number of babies with given `name`
- `prop`: proportion (or frequency)


In order to make a timeline plot with the trend for your name, 
it's better to subset `babynames` with just the relevant values:
```{r}
me <- 'Gaston'   # change this value!
mynames <- subset(babynames, name == me)
```


### Timeline Plot

Once we have the required subset, we can plot a timeline trend: 
```{r}
ggplot(data = mynames, aes(x = year, y = n)) +
  geom_line(size = 1)
```

Because some names appear in both males and females, we can add color based on
sex:
```{r}
ggplot(data = mynames, aes(x = year, y = n)) +
  geom_line(size = 1, aes(color = sex))
```


- Find how many people have been named like you
- Find the year in which your name was the most popular

```{r}
# number of people named like you
sum(mynames$n)

# most popular year
mynames$year[which.max(mynames$n)]
```

My name was more popular than ever in `r mynames$year[which.max(mynames$n)]`.
So far, `r sum(mynames$n)` have been named `r me`.

-----


## Getting Started with Shiny

__"shiny"__ is an R package that does two things:

1. Creates __Reactive__ objects
2. Build HTML web pages

Perhaps the most popular use of shiny is for creating Web Apps, commonly known
as "shiny apps". You can take a look at some examples of shiny apps at:
[http://shiny.rstudio.com/gallery](http://shiny.rstudio.com/gallery) 

Right now we won't be using shiny to create a web app (we'll do that at the
end of the course). Instead, we are going to see how to use shiny to work 
with __reactive objects__ and create a document with dynamic plots (not 
exactly interactive plots, but close enough).


### Shiny Interactive Document

We'll generate an interactive document editing an Rmarkdown (.Rmd) file and 
using shiny. To open an interactive document in RStudio:

1. Go to __File__, click __New File__, then choose __R Markdown...__
2. Select __Shiny__, then __Shiny Document (default)__
3. Delete everything under the second `---`

In the Rmd file make sure you have the following _yaml_ header (you may include 
an optional title):

```bash
---
runtime: shiny
output: html_document
---

## What's in a name?
```

### Reactivity 101

There are two main types of reactive objects in shiny:

- __Widgets__: let users set a value by typing, clicking, scrolling, etc.
- __Rendered__: respond whenever a widget value changes

The first thing to do is to include a widget that allows you to type some text 
(your name). Use the __textInput()__ widget inside a code chunk:

~~~ 
```r
# change the value to your name
textInput(inputId = "name", label = "Name:", value = "Gaston")
```
~~~

Your widget saves a value in R that you can call with `input$<inputId>`.
For example, let's say you have a `textInput` widget with the following call:
`textInput(inputId = "name", label = "Name:", value = "Gaston")`.
You would call the value of this widget with `input$name`.

The nice thing about widgets is that a its value changes whenever a user changes the widget.

You have to be very careful with widgets: You CANNOT call a widget value with 
a normal R function. For instance,  if you try to count the number of characters
in `input$name` doing something like this `nchar(input$name)`, Shiny won't let 
you.

To create a widget you'll have to use widget functions. The table below lists
the available widgets functions in shiny:

| Function        | Widget                                     |
| --------------- | ------------------------------------------ |
| `actionButton()`   | Action Button                              |
| `checkboxGroupIn()` | Group of checkboxes                        |
| `checkboxInput()`   | Single checkbox                            |
| `dateInput()`       | Calendar to aid date selection             |
| `dateRangeInput()`  | Pair of calendars for selecting date range |
| `fileInput()`       | File upload control wizard                 |
| `helpText()`        | Help text to accompany other widgets       |
| `numericInput()`    | Field to enter numbers                     |
| `radioButtons()`    | Set of radio buttons                       |
| `selectInput()`     | Box with choices to select from            |
| `sliderInput()`     | Slide bar                                  |
| `submitButton()`    | Submit button                              |
| `textInput()`       | Field to enter text                        |

More info about widgets at: [Shiny Widgets Gallery](http://shiny.rstudio.com/gallery/widget-gallery.html)



### Reactive expressions

The way to manipulate a reactive value (like `input$name`) is to use it 
within a __render function__. 
For instance, to be able to display your name, you need to pass `input$name` 
to the function `renderText()` like so:
```r
`renderText(input$name)`
```

Or include `renderText(nchar(input$name))` in an inline code chunk.

Below is a list with the available render functions in Shiny:

| Function            | Creates       |
| ------------------- | ----------------- |
| `renderDataTable()` |	An interactive table |
| `renderImage()`     |	An image (saved as a link to a source file) |
| `renderPlot()`      |	A plot            |
| `renderPrint()`     |	A code block of printed output |
| `renderTable()`     |	A table            |
| `renderText()`      |	A character string |


### Your Turn

Work on the file `lab04-shiny-doc-raw.Rmd` (located in the `docs/` directory 
of the github repo of the course) to create a shiny document. You
should be able to input your name, and then see the produced plot with the 
line trend of your name, what was the most popular year, and how many people
have been named like you.

If you get stuck on a specific part, try to solve it first, ask the GSI, 
ask your neighbor, or google it. If you give up, then take a peek at the 
solutions in `lab04-shiny-doc-final.Rmd` (located also in the `docs/`
directory of the course github's repo)

-----

