---
title: "Rmd Code Chunks Demo"
author: "Shamindra Shrotriya"
date: "October 12, 2016"
output: html_document
---

## Background

This is a quick interactive Rmd file used to explain the various `code chunk`
options availible in R Markdown.

It was inspired by the following great resources (please read them):

*  [Karl Borman's Rmd tutorial](http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html)
*  [Yihue Xie's Rmd Chunk Options page](http://yihui.name/knitr/options/). 
    *  _Fun Fact:_ Yihue created `knitr`!

### R Markdown Code Chunks

The general format of an Rmd code chunk is as shown below

```
{r chunk_name, chunk_options}
```

### Option 1: `include=FALSE`

Here we just set up some knitr options we want the code to be 
_evaluated_ but **neither code or output to be** _displayed_:

```{r include_false_demo, include=FALSE}
library(knitr)
opts_knit$set(root.dir = '~/Downloads')
```

The code that was hidden in the final knit document was:

```
opts_knit$set(root.dir = '~/Downloads')
```

### Option 2: `results='hide'`

Here we just set up some knitr options we want the code to be 
_evaluated_ but **the output not to be** _displayed_:

```{r results_hide_demo, results='hide'}
summary(cars)
```

### Option 3: `echo=FALSE`

Here we just set up some knitr options we want the code to be 
_evaluated_ but **only the output to be** _displayed_:

```{r echo_false_demo, echo=FALSE}
plot(cars$speed)
```

### Option 4: `eval=FALSE`

Here we just set up some knitr options we want the code to 
**not be evaluated** but **only the code to be** _displayed_:

```{r eval_false_demo, eval=FALSE}
min(cars$dist)
```

We didn't get the minimum value of the `distance` variable in `cars` 
dataframe displayed as expected here, just the R code command

### Option 5: `cache=TRUE`

Here if the code chunk remains unchanged between 'knitting', then it
won't be re-run in the next knit i.e. RStudio will cache (save) the output
from the previous knit and not re-run.

This is used to **speed up the R Markdown development process** by reducing
the time for each subsequent knit process.

```{r cache_true_demo, cache=TRUE}
max(cars$dist)
```

### Option 6: `warning=FALSE, message=FALSE`

Here `warnings` and `messages` of `R` output are suppressed. By default they
are not. In general you should not require this but there are some special 
cases where it is useful.

```{r warning_message_false_demo, warning=FALSE, message=FALSE}
library(knitr)
```

### Option 7: `Preferred for STAT133`

In `STAT133` we (i.e. Instructor and GSIs) want to see your code and output
displayed in the knit html. So there should be no need to have `echo=FALSE`
or `results='hide'` or `eval=FALSE`

```
{r chunk_name, cache=TRUE}
```